\section{Discussion}
\label{sec:discussion}

\paragraph{Relation to active inference.}
Equation~\ref{eq:F} is a faithful instantiation of the variational
free energy with an amortized accuracy decoder. The JKO step
coincides with the Wasserstein-lifted variational update, which in
the convex regime would yield the unique posterior. Our $F$ is
non-convex through the JEPA term ; we observe in practice that 10
JKO steps suffice to produce informative $\rho_\tau \to \rho_{\tau+1}$
pairs without instability. The non-convexity is structural (a neural
decoder composed with a Wasserstein metric cannot be made convex
without collapsing the model), so we treat it as a feature to be
managed (via step-size and step-count hyperparameters) rather than a
bug to be eliminated. The empirical stability we observe across
50k queries at two scales (10k, 50k) suggests that the functional is
well-posed in the basin of attraction around the prior, even if
global uniqueness is not guaranteed.

\paragraph{What this means for the free-energy community.}
Most active-inference implementations either (a) restrict to
tractable posterior families (Gaussian, factorized categorical)
that admit closed-form updates or (b) rely on message-passing
approximations whose convergence is only local. \textsc{QueryConditionedF}
offers a third path: use a neural amortization for the accuracy term,
pay the price of non-convexity, and recover tractability through the
Wasserstein geometry of JKO. This is a small but concrete data point
that structured-latent active inference with pretrained decoders is
feasible at realistic scale (50k queries, 0.65M--3.4M encoder
parameters). The R3 kill-switch methodology --- ranking pilot
architectures, freezing the top-$k$ by a pre-registered threshold,
then scaling --- is offered as a reusable protocol for other
free-energy benchmarks that face compute constraints.

\paragraph{Limitations.}
\textit{Semantic bin mapping.}
The semantic bin mapping currently uses a stable lemma hash rather
than a true WordNet-fr or KMeans-MiniLM clustering --- an absolute
identity choice that does not affect the simplex structure but does
limit the cognitive interpretability of individual bins. A swap-in
KMeans-MiniLM replacement is straightforward and would change per-bin
identities without changing the aggregate KL ranking.

\textit{Empirical ecological validity.}
Alignment with human reading-time or fMRI data is out of scope for
this workshop submission; all of our validation is intrinsic to the
simplex dynamics (KL divergence on held-out test split, MAPE on
predicted delta, routing hit-at-5). Extrinsic validation against
psycholinguistic behaviour is planned for the full paper.

\textit{Joint training of $f_\theta$ and $g_{\mathrm{JEPA}}$.}
We also defer joint self-supervised refinement of $f_\theta$ and
$g_{\mathrm{JEPA}}$ (phase B) to a companion full paper. In the
present work, the decoder is frozen during encoder training, which
means encoder rankings depend on the specific heuristic labeler.
Ablating the labeler (e.g., replacing Lexique.org with SUBTLEX-fr)
would test the robustness of the ranking and is a concrete follow-up.

\textit{Corpus composition.}
The 50k hybrid corpus weights psycholinguistic seeds (B, 20\,\%),
Wikipedia generalist (C, 40\,\%), and Qwen3.5-35B synthetic (D, 40\,\%).
Scaling to 500k or replacing D with a distilled-model-free generator
is an open question. The R3 kill-switch logic carries over, but
the 15\,\% gap threshold may need re-calibration at that scale.

\paragraph{Future work.}
Phase B will train $f_\theta$ and $g_{\mathrm{JEPA}}$ jointly via
differentiable JKO with optional contrastive variants of the JEPA
loss; that line of work targets ICLR. Beyond phase B, we see three
concrete directions: (i) extending the simplex decomposition from
four species to finer-grained Levelt-Baddeley components
(articulatory loop, visuo-spatial sketchpad); (ii) coupling
\textsc{QueryConditionedF} to the Nerve
Protocol~\citep{saillant2026nervewml} to test whether text-driven
dynamics preserve the cross-modal invariants B-1 / B-2 / B-3
measured in \texttt{bouba\_sens}; and (iii) ablating the heuristic
labeler against a learned labeler obtained by distilling a large
language model's token-to-species posterior, which would make the
functional fully end-to-end differentiable.

\paragraph{Reproducibility statement.}
All code, configs, and cached oracle runs are available at
\url{https://github.com/hypneum-lab/kiki-flow-research} (tag to be
assigned at submission time). Encoder training uses JAX on
RTX 4090; the oracle runs use MLX on Apple M3 Ultra. SHA256 hashes
of every \texttt{.safetensors} cache entry are logged in the
submission bundle, and the R3 kill-switch decision is recorded in
\texttt{docs/superpowers/specs/2026-04-19-query-conditioned-f-design.md}.
