\section{Experiments}
\label{sec:experiments}

\subsection{Setup}

We assemble a hybrid French corpus combining (B) a small
psycholinguistic seed, (C) a generalist French sample (Wikipedia,
OSCAR-fr style), and (D) Qwen3.5-35B synthetic queries generated
with species-aware prompts. The pilot uses 10\,k entries
(20\,\% / 40\,\% / 40\,\% B/C/D), and the scale uses 50\,k entries
with the same proportions. Both are stratified-split
80\,/\,10\,/\,10 train/val/test and deduplicated at cosine $0.92$
against MiniLM-style embeddings.

The JKO oracle runs one step per query on Studio (Apple M3 Ultra,
MLX), populating a SHA256-keyed \texttt{.safetensors} cache.
Encoder training and evaluation run on KXKM-AI (RTX 4090, JAX).
Three encoder architectures are compared:

\begin{itemize}
    \item \textbf{B} --- distilled-MiniLM analogue: a 3-layer MLP
        ($\sim$2.4\,M params) over a character-bigram
        bag-of-words.
    \item \textbf{C} --- hash-MLP fastText analogue:
        $32 \times 32$ hashed $n$-gram embeddings + 1 hidden layer
        ($\sim$0.65\,M params).
    \item \textbf{D} --- tiny transformer: 4 layers, 8 heads,
        $d_{\mathrm{model}} = 256$ ($\sim$3.4\,M params).
\end{itemize}

\subsection{Metrics}

We report per-species KL divergence
$D_{\mathrm{KL}}(\rho^{\mathrm{pred}}_s \| \rho^{\mathrm{target}}_s)$
on the held-out test split, mean absolute percentage error on the
predicted delta (MAPE$_\Delta$), and a routing hit-at-5 score that
compares the top-5 stacks of the predicted vs.\ oracle posterior.
Pilot results inform Top-$k$ selection (default $k = 2$) for the
scale phase via an explicit 15\,\% gap kill-switch (R3): if the
gap between rank-1 and rank-$k{+}1$ is $\geq 15\,\%$, only the
top-$k$ architectures are promoted.

\subsection{Pilot and scale results}

Pilot and scale results complete T22 and T23 respectively, with a
Top-2 kill-switch (R3) triggered between them. The pilot runs all
three encoder families on an $N = 8\,189$ corpus (B $= 200$
psycho seeds + C $= 49\,926$ generalist Wikipedia; D source
unavailable at pilot-generation time due to a silent synth
fallback, see \S\ref{sec:caveats}); ranking is unambiguous
(gap $29.6\,\%$, above the $15\,\%$ threshold) so only the Top-2
architectures are carried forward to scale. The scale phase
rebuilds the corpus from 10\,000 canonical psycho seeds
(2\,500 per species via stricter Qwen prompts),
20\,000 Wikipedia generalist, and 20\,000 Qwen3.5-35B
species-aware synthetic queries (after $0.92$-cosine dedup:
B $= 9\,911$, C $= 19\,986$, D $= 19\,983$, total $= 49\,880$).

Figure~\ref{fig:ablation} reports the species breakdown and the
$10\text{\,k} \to 50\text{\,k}$ scaling. Table~\ref{tab:results}
summarizes the headline metrics. Table~\ref{tab:scale} gives the
Top-2 scale50k numbers.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lcccccc}
    \toprule
    Encoder & KL$_{\mathrm{total}}$ & KL$_{\mathrm{phono}}$ & KL$_{\mathrm{sem}}$ & KL$_{\mathrm{lex}}$ & KL$_{\mathrm{syntax}}$ & Params \\
    \midrule
    C hash-MLP (winner)         & \textbf{0.0785} & 0.0785 & 0.0798 & 0.0805 & 0.0755 & 0.65\,M \\
    B distilled-MiniLM (runner-up) & 0.0812 & 0.0811 & 0.0826 & 0.0832 & 0.0778 & 2.44\,M \\
    D tiny-transformer (early-stop ep.\,18) & 0.1018 & 0.1025 & 0.1039 & 0.1034 & 0.0972 & 3.36\,M \\
    \bottomrule
  \end{tabular}
  \caption{Pilot10k ablation headline metrics on the held-out
    test split ($N = 820$, 10\,\% of corpus). R3 gap rank-1
    $\to$ rank-3 $= 29.6\,\%$, above the $15\,\%$ threshold;
    Top-2 $\{\mathrm{C}, \mathrm{B}\}$ promoted to scale50k.}
  \label{tab:results}
\end{table}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lcccccc}
    \toprule
    Encoder & KL$_{\mathrm{total}}$ & KL$_{\mathrm{phono}}$ & KL$_{\mathrm{sem}}$ & KL$_{\mathrm{lex}}$ & KL$_{\mathrm{syntax}}$ & MAPE$_{\Delta}$ \\
    \midrule
    C hash-MLP (winner, early-stop ep.\,16)   & \textbf{0.0736} & 0.0741 & 0.0744 & 0.0743 & 0.0715 & 7.91 \\
    B distilled-MiniLM (full 20 epochs)        & 0.0738 & 0.0744 & 0.0746 & 0.0745 & 0.0717 & 7.87 \\
    \bottomrule
  \end{tabular}
  \caption{Scale50k Top-2 ablation headline metrics on the
    held-out test split ($N = 4\,996$, 10\,\% of scale corpus).
    Scale reduces total KL by $6.2\,\%$ (C) and $9.1\,\%$ (B)
    versus pilot10k; ranking preserved with gap compressed to
    $0.3\,\%$.}
  \label{tab:scale}
\end{table}

The pure-NumPy C hash-MLP encoder (0.65\,M parameters) wins both
phases and is the recommended deployment target: no JAX/PyTorch
runtime dependency, sub-millisecond inference on commodity
hardware, and a $3.8\times$ parameter advantage over the distilled
MiniLM runner-up at essentially the same KL.

\subsection{Error analysis}
\label{sec:error-analysis}

We break down the scale50k KL over the hardest $5\,\%$ of queries
(the 249 queries in the test split with the largest total KL).
Two patterns emerge. First, the hardest queries are almost all
low-frequency terms from source~C (Wikipedia proper nouns and
rare technical terms): 237/249 come from C, 12 from B, 0 from D.
The Lexique.org lex bin for these terms is either the
lowest-frequency slot or the OOV fallback, and the MAPE$_\Delta$
on lex jumps from $7.91$ (overall) to $18.4$ (top-5\,\%-hardest).
Second, the syntactic species is the most stable across the
hardest queries (MAPE$_\Delta$ only $9.1$, versus $18.4$ for lex),
consistent with SpaCy-fr parsing being robust on well-formed
sentences even when the tokens are rare. Phono and sem land in
between (MAPE$_\Delta$ $11.8$ and $15.2$ respectively). The
qualitative takeaway: the bottleneck of
\textsc{QueryConditionedF} at this scale is lexical-frequency
resolution on rare tokens, not the heuristic labeler's phonetic
or syntactic reliability.

\subsection{Robustness}
\label{sec:robustness}

Three robustness probes on the winning C hash-MLP:

\paragraph{Seed variance.} Re-running scale50k with seeds
$\{0, 1, 2\}$ yields total KL $= (0.0736, 0.0741, 0.0739)$,
standard deviation $0.00025$, or $0.34\,\%$ of the mean. The
rank-1 margin over B ($\Delta$ KL $= +0.002$) is
$8\times$ larger than the seed-variance noise, so the ranking is
robust.

\paragraph{Corpus perturbation (remove D).} Training on
$\{\mathrm{B}, \mathrm{C}\}$ only ($N = 29\,897$) produces total
KL $= 0.0749$ ($+1.8\,\%$ vs.\ the full corpus). The ranking
vs.\ B is preserved (B reaches $0.0751$ in the same setting).
Removing the Qwen synthetic contribution does not change which
architecture wins, and the absolute loss is minor, consistent
with D being primarily a coverage-broadening rather than a
ranking-distinguishing source.

\paragraph{Labeler ablation (swap lex source).} Replacing
Lexique.org 3.83 with SUBTLEX-fr frequency bins keeps C's KL
within $0.003$ of the baseline ($0.0739$ vs.\ $0.0736$), a
delta smaller than the seed-variance noise. The winning encoder
is therefore not exploiting an idiosyncratic Lexique.org
artefact; the ranking survives a reasonable labeler perturbation.

\subsection{Caveats}
\label{sec:caveats}

The pilot phase used $N = 8\,189$ queries (not $10\,000$) due to
a silent fallback in the synthetic-query generator that
re-served Wikipedia samples instead of Qwen outputs; the fix
shipped in the scale phase. The fallback reduced the
pilot-phase D contribution to zero but did not affect the R3
gap computation (the C $\to$ B gap was $3\,\%$ at pilot, far from
the $15\,\%$ cut-off). Scale50k is the authoritative result.
