\section{Introduction}
\label{sec:intro}

Text encoders that produce dense embeddings have become a foundational
primitive in modern machine learning, yet the question of how an
embedding should \emph{act on} a structured cognitive state has
received less attention. We address one instance of this problem:
coupling a 384-dimensional French sentence embedding to a Wasserstein
gradient flow over a four-species psycholinguistic simplex inspired
by the Levelt-Baddeley working-memory model.

\paragraph{Positioning.} Three adjacent lines of work approach
text-to-latent coupling from different angles. Classical JEPA
\citep{lecun2022} trains a context encoder to predict target
embeddings in a shared representation space but makes no structural
commitment about that space. Amortized variational inference with
continuous normalizing flows (e.g., \citealp{chen2018neural}) gives
principled dynamics but operates on unconstrained $\mathbb{R}^d$
rather than on a bounded simplex. Active-inference
\citep{friston2010,parr2017} provides the right variational
scaffold --- accuracy plus complexity --- but its computational
instantiations typically use Gaussian or factorized categorical
posteriors, not probability-simplex latents driven by an
external query. None of the three, to our knowledge, simultaneously
(a) bounds the latent to a probability simplex, (b) conditions
the dynamics on an external text query, and (c) preserves the
active-inference free-energy decomposition. \textsc{QueryConditionedF}
is the first functional we are aware of that does all three.

\paragraph{The query-agnostic baseline problem.} Two existing
free-energy functionals in our codebase, \texttt{T1FreeEnergy} and
\texttt{T2FreeEnergy}, are query-agnostic. Their dynamics depend
only on the current state, the species coupling matrix, and a fixed
prior. Using either as the oracle for a text-conditioned bridge
surrogate produces \texttt{(state\_pre, state\_post)} pairs whose
post-state is independent of the query, defeating the purpose of
training a text-aware downstream model. \textsc{QueryConditionedF}
is the minimal modification that makes the oracle \emph{react}
to the query.

\paragraph{Why this is non-trivial.} Three difficulties arise.
First, the JEPA accuracy term is non-convex in $\rho$, so the JKO
step is no longer guaranteed to converge to a unique minimum, and
we observe empirically that step-size and number of steps both
matter. Second, the state is a product of four 32-bin simplices
rather than a flat $\mathbb{R}^d$ --- the gradient must be projected
back onto the simplex after each update, which we achieve via a
softplus-normalize scheme detailed in \S\ref{sec:query-conditioned-f}.
Third, the decoder $g_{\mathrm{JEPA}}$ requires pre-training, yet no
ground-truth species labels exist for French text; we handle this
cold-start via a deterministic heuristic labeler built from
SpaCy-fr, Lexique.org, and \texttt{phonemizer}.

\paragraph{Proposal.} We propose \textsc{QueryConditionedF}, a
third free-energy subclass whose accuracy term is parameterized
by a JEPA-style decoder
$g_{\mathrm{JEPA}}: \Delta^{128} \to \mathbb{R}^{384}$ that
reconstructs the query embedding from the simplex state.
The complexity term is the standard per-species KL to a
Levelt-Baddeley prior, and a lightweight species-coupling term
carries over the inter-code interactions of \texttt{T2FreeEnergy}.
The decoder is pre-trained offline against heuristic labels
obtained from a deterministic French NLP pipeline (SpaCy-fr,
phonemizer, Lexique.org) --- sidestepping the need for ground-truth
species annotations.

\paragraph{Contributions.} (i) A principled formulation of
\textsc{QueryConditionedF} with analytical gradients for the
complexity and coupling terms and JAX autodiff for the JEPA term;
(ii) a reproducible heuristic French labeler producing 32-simplex
targets per species; (iii) an empirical ablation of three encoder
architectures over a 50k hybrid French corpus with an explicit
R3 kill-switch (15\% gap threshold) between pilot and scale phases,
demonstrating that text conditioning shifts the post-state
distribution in a predictable, measurable way. A companion full
paper (in preparation) extends this design to joint self-supervised
training of the encoder and decoder via differentiable JKO and
reports on a biological-bridge extension using the Nerve Protocol
\citep{saillant2026nervewml}.

\paragraph{Reproducibility and pre-registration.} All results are
reproducible from the tagged repository
(\url{https://github.com/hypneum-lab/kiki-flow-research}), with
SHA256-keyed \texttt{.safetensors} caches for every oracle run.
The R3 kill-switch threshold (15\% relative gap rank-1$\to$rank-3)
was fixed before the pilot and is reported verbatim in \S\ref{sec:experiments}.
